Cluster Difference Imaging Photometric Survey. III. An Open Cluster With a 250\,pc Halo

### 2021-01-29 cluster_paper/ms.tex
paper updates

@@

\title{
  Cluster Difference Imaging Photometric Survey. III.
  The Coeval Halo and Core of NGC\,2516
}

@@

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations (``halos'') surrounding nearby open clusters.  The stars in these
  halos could have escaped from the cores of the cluster, or they
  could be remnants of a larger and lower-density star formation
  complex that has since dispersed.  They could also be field
  stars---false positives given by overly aggressive clustering algorithms.
  In this study, we focus on a single reported 250\,pc-long halo
  around the $\approx$120\,Myr open cluster NGC\,2516.  The classical tidal radius of
  this cluster is $\sim$XX\,pc.  Combining photometry from Gaia,
  rotation periods from TESS, and lithium measurements from
  the Gaia-ESO and GALAH surveys, we find that: i) the halo of
  NGC\,2516 is real and coeval with the core; ii) at fixed stellar
  mass and age, rapidly rotating stars have greater lithium abundances
  than slowly rotating stars; and iii) given Gaia-selected halo
  members, the contamination rate of field stars in the halo (XX\%) is
  larger than that in the core (YY\%).  This work expands the set of
  confirmed NGC\,2516 members by a factor of $\approx$2, and
  quantifies the degree to which Gaia-based analyses can currently
  identify dispersed stellar populations against a background of field
  star contaminants.  Implications for spectroscopic survey targeting, open cluster dispersal, and planet searches around young
  stars are discussed.
\end{abstract}

@@

Recent clustering studies using Gaia have begun to report the
identification of structures that could correspond to low-density halos
of stars that may have evaporated from open clusters (e.g., Kounkel+19,
Kounkel+20, Meingast+21).  However, different clustering methods on the
Gaia data tend to give different results (CITE: Hunt \& Reffert 2020).
Employing say Gaussian Mixture Modelling, or any analogous method that
requires ``clusters'' to be ellipses in phase-space (position or
velocity) unsurpsingly yields open clusters that are roughly elliptical.
Some unsupervised clustering methods lead to comparable ``roughly
elliptical'' results \citep[][hereafter
\citetalias{cantatgaudin_gaia_2018}]{cantatgaudin_gaia_2018}.  Other
unsupervised methods such as HDBScan (e.g., \citep[][hereafter
\citetalias{kounkel_untangling_2019}]{kounkel_untangling_2019}) have
been found to yield additional structures, particularly in lower density
regions such as the Psc-Eri stream (CITE: Meingast+18, Curtis+19,
Newton+21), but also more generally around many open clusters
\citep{kounkel_untangling_2019}.  Unsupervised approaches that
incorporate physical constraints like imposing a maximum velocity
dispersion on putative members yield similar results, potentially with
higher purity (Meingast+21).

@@

As part of a broader project of identifying a large and clean sample of
young stars for a transit search, we focus in his paper on a rather
modest question: in just a single rich southern open cluster, is the
cluster halo coeval with the core?  The cluster we chose for this
analysis was NGC\,2516, since it was young ($\sim$120\,Myr) and close
($d\approx400$\,pc) enough to facilitate rotation measurements using
TESS, and some degree of spectroscopic analysis.  We want to know: is
the halo real? To what extent can we use Gaia alone to reliably identify
age-dated needles in the haystack of boring field stars?  And more
generally, what are the implications for the evolution of open clusters
if they do have halos?

@@

We chose the stars to focus on based on previously reported
memberships of NGC\,2516, available in the literature.  While a number
of pre-Gaia catalogs were available, the purity and accuracy of the
Gaia-derived results are expected to be the current state of the art.
We therefore adopted what we viewed as the most interesting clustering
samples to compare: those of \citet{cantatgaudin_gaia_2018}
(\citetalias{cantatgaudin_gaia_2018}) and
\citet{kounkel_untangling_2019}
(\citetalias{kounkel_untangling_2019})\footnote{A third sample was
also recently reported by \citet{meingast_2021}. Although the results
of this study were made publicly available too late to be included in
our analysis,  1577 of the 1860 \citet{meingast_2021} sources (85\%)
were included in the \citet{kounkel_untangling_2019} sample.  A
complete set of \cn members could very well include some of the newly
reported candidate members from \citet{meingast_2021}.}. While we
could have performed our own clustering analysis based on the Gaia
data, such an effort would be replicating the work of these
investigators. And why would our clustering methods be better than
theirs? We opt instead to use their studies as starting points.

@@

%TODO calculate these physical sizes carefully
Figure~\ref{fig:gaia6d} shows the cluster members reported by each
study.  The \citetalias{cantatgaudin_gaia_2018} members are all within
a few degrees of the cluster center, while the
\citetalias{kounkel_untangling_2019} members span tens of degrees.  In
physical units, this corresponds to a size difference of XX\,pc versus
YY\,pc.  This difference is somewhat tautological, because
\citetalias{cantatgaudin_gaia_2018} did not extend their search for
members out to tens of degrees from the cluster center, since their
clustering algorithm relied on the contrast between cluster and field
stars in the astrometric space of ($\mu_{\alpha'}, \mu_\delta, \pi$).
Nonetheless, the membership catalog of \citetalias{cantatgaudin_gaia_2018}
echoes that of many previous investigators (CITE, CITE, CITE), and is
consistent with the general {\it visual} impression that one has when
looking at \cn visually: it seems to be at most {\bf X$^\circ$}, and
not a big stream.
This discrepancy begs the question.  What is the true structure of
\cn?  Are the core and halo truly coeval?

To answer these questions, we adopt the
\citetalias{cantatgaudin_gaia_2018} stars as candidate ``core''
members and the \citetalias{kounkel_untangling_2019} stars as
candidate ``halo'' members.  If a star is in both lists, it is a
``core'' member.  1091 of the 1106 core members are in both lists.
This leaves 1912 candidate members in the halo.
Concatenating the two lists yields 3018 candidate members.

In this work, we also define a set of nearby field stars in the
``neighborhood'' of \cn.  Based on the observed distribution of halo
members, we draw these stars randomly from the following ``cube'' in
right ascension, declination, and parallax:
\begin{align}
  $\alpha$\,[{\rm deg}] &\in [108, 132], \\
  $\delta$\,[{\rm deg}] &\in [-76, -45], \\
  $\pi$\,[{\rm mas}] &\in [1.5, 4.0].
\end{align}
We imposed a magnitude limit of $G=19$\,mag, and ran the queries
using the \texttt{astroquery.gaia} module (CITE).
% TODO: why?
We allowed the number of stars in the comparison sample to exceed that
in the cluster sample by a factor of a few, to ensure broad sampling
of stellar masses and evolutionary states.
We also required the comparison sample to not overlap with the cluster
sample, which led to the omission of 1.1\% of the stars drawn over the
volume noted above.

@@

		\subfloat{
			\includegraphics[width=0.75\textwidth]{f3b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf The core and halo of \cn in the space of rotation
    period and Gaia color.}
    The {\it top} plot shows periods against a linear scale, while the
    {\it bottom}  shows them against a logarithmic scale.
		\label{fig:rot}
	}
\end{figure*}

@@

We then measured the rotation period of the resulting light curve.  We
used the the aperture radius that, based on theoretical expectations,
was expected to give the optimal balance between light from the target
and background-light (CITE Sullivan15).  This typically resulted in an
aperture radius of either 1 or 1.5 pixels.  To measure the periods, we
used the periodogram implementations in \texttt{astrobase}, in
particular the Stellingwerf PDM periodogram (CITE), along with the
more traditional Lomb-Scargle (CITE).  We recorded the top five
periodogram peaks from each method, and their corresponding powers.
Finally, as a check on crowding, we also recorded the number of stars
within the aperture of equal brightness, and of brightness with 1.25
and 2.5 TESS magnitudes of the target star.

Figure~\ref{fig:rot} shows the resulting rotation periods.  The points
on this plot are light curves for which the peak Lomb Scargle
periodogram period was below 15 days, had power exceeding 0.08, and
had at most one equal-brightness companion within the aperture.  These
selection criteria are entirely heuristic.  
But splitting the sample into the ``core'' and the ``halo'', they
suggest that the halo is real.

@@

How far away from the core, in position and velocity space, does the
halo actually extend?  We can explore this by crossmatching the
rotator sample against the Gaia data.  Figure~\ref{fig:gaia6d_x_rotn}
shows the result.

@@

Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), it seems that the rapid rotators tend to show elevated
lithium equivalent widths.  Similar trends have been previously noted
in the Pleiades by \citet{soderblom_evolution_1993} and
\citet{bouvier_pleiades_lirot_2018}, in the Psc-Eri stream by
\citet{arancibia_2020}, and in M\,35 by \citet{jeffries_m35_li_2020}.

% TODO: this paragraph more or less parrots the processes described by
% Bouvier+20. are there others? is there a better way to describe
% these?
The trend contradicts what might be the naive expectation that rapidly
rotating stars should show more vigorous convection, leading to
earlier destruction of photospheric Li.  Possible physical
explanations include processes both internal and external to the star
(CITE see the recent review by Bouvier+20).  Internal processes would
be tied to the interplay between surface rotation, differential
rotation in the interior, and the convective mixing efficiency (e.g.,
CITE Siess + Livio 1997, Baraffe+2017).  Internal processes could also
include the possibility that stronger magnetic fields in the stellar
interior inhibit convection (READ e.g., Ventura+98, Chabrier+07,
Somers + Pinsonneault 2014).  An external process that could also be
important is the effect of star-disk magnetic braking during the PMS
phase (CITE: magnetic braking).  Longer disk lifetimes would lead to
the star's outer convective zone being ``locked'' for longer while the
radiative core continues contracting.  The resulting differential
rotation and rotational mixing could drive the lithium depletion
(CITE: Bouvier 08, Eggenberger+12).

@@

  \subsection{What is the contamination fraction in the halo? Does it change vs. location?}
  \subsection{Are the ``very slow rotators'' bonafide members?}
  Probably not. They are not isotropically distributed around the
  cluster.
  \subsection{How did the halo form?}
  \subsection{Mass differences between center and outer reaches?}
  \subsection{Fast rotators: are we going faster than other clusters?}

@@

@@

%
% TODO: the following two paragraphs, when quoted, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citetalias{cantatgaudin_gaia_2018} applied a procedure that ultimately
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by Dias.  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.
When we selected candidate \cn members from the results of
\citetalias{cantatgaudin_gaia_2018}, we opted to include all candidate members
with reported membership probability exceeding 10\%.
While this seems {\it a priori} low, our results ({\bf SECTION XXX})
show that this ``membership probabilty'' severely underestimates the
purity of the \citetalias{cantatgaudin_gaia_2018} sample for \cn. 
%TODO: quantify
Their false positive rate across the sample is more like 1-5\%.

\citetalias{kounkel_untangling_2019}
applied a different unsupervised
clustering method to the 5-dimensional Gaia DR2 data (omitting radial
velocities, due to their sparsity).  Their selection function (see
their Section~2) yielded $\approx 2\times 10^7$ stars, mostly within
$\approx 1$\,kpc and typically with $G<18$\,mag.  Their clustering
algorithm was the ``hierarchical density-based spatial clustering of
applications with noise'' (HDBSCAN, CITE McInnes17).  The classical
DBSCAN algorithm ``identifies clusters as overdensities in a
multi-dimensional space in which the number of sources exceeds the
required minimum number of points within a neighborhood of a
particular linking length $\epsilon$.  HDBSCAN does not depend on
$\epsilon$; instead it condenses the minimum spanning tree by pruning
off the nodes that do not meet the minimum number of sources in a
cluster and reanalyzing the nodes that do. Depending on the chosen
algorithm, it would then either find the most persistent structure
(through the excess of mass method), or return clusters as the leaves
of the tree (which results in somewhat more homogeneous clusters). In
both cases it is more effective at finding structures of varying
densities in a given data set than DBSCAN.''
`` The two main parameters that control HDBSCAN are the number of
sources in a cluster and the number of samples. The former is the
parameter that rejects groupings that are too small; the latter sets
the threshold of how conservative the algorithm is in its
considerations of the background noise (even if the resulting noisy
groupings do meet the minimum cluster size).  By default, the sample
size is set to the same value as the cluster size, but it is possible
to adjust them separately.''
Regarding membership proabilities,
\citetalias{kounkel_untangling_2019} did not report continuous membership
probabilities, instead opting for the binary ``member'' or ``not''.

@@

There are many methods for determining the coefficients of the linear
model, after the full set of eigenvectors (plus optionally
``sytematic'' vectors) has been asssembled.  We explored two: ordinary
least squares, and ridge regression. Ridge regression is the same as
ordinary least squares, except it includes an L2 norm with a regularization
coefficient. The regularization coefficient that best applied for any
given target light curve was solved for using a cross-validation grid
search, using \texttt{sklearn.linear\_modelRidgeCV} (CITE).

Each target light curve was mean-subtracted and normalized by its
standard deviation, as were the eigenvectors. The linear problem was
then solved, and the light curve was reconstructed by re-adding the
original mean, and re-multiplying by the standard deviation to ensure
that the variance of the light curve did not change.

We found, somewhat to our surprise, that the choice of using ordinary
least squares versus ridge regression did not seem to significantly
affect the resulting light curves. In other words, the inclusion (or
lack thereof) of a regularization did not strongly alter the
best-fitting coefficients.

@@

>>>>

\newcommand{\nkinematic}{3{,}018\ } % "fullfaint" kinematic sample, core+halo. from any call of earhat.helpers._get_fullfaint_dataframes
\newcommand{\nautorot}{1{,}276\ } % drivers.plot_full_kinematics_X_rotation; 2021.01.29

@@

\title{
  Cluster Difference Imaging Photometric Survey. III.
  %The Coeval Core \& Halo of NGC\,2516
  %The Halo of NGC\,2516 Extends To At Least 250\,pc
  An Open Cluster With a 250\,pc Halo
  %NGC\,2516  250\,pc
}

@@

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations (``halos'') surrounding nearby open clusters.  The stars in these
  halos could have escaped from the cores of the cluster, or they
  could be remnants of a star formation
  complex that has since dispersed.  They could also be field
  stars---false positives originating from overly aggressive clustering algorithms.
  In this study, we focus on a single reported 250\,pc-long halo
  around the $\approx$120\,Myr open cluster NGC\,2516.  The classical tidal radius of
  this cluster is $\sim$XX\,pc.  Combining photometry from Gaia,
  rotation periods from TESS, and lithium measurements from
  the Gaia-ESO and GALAH surveys, we find that the halo of
  NGC\,2516 is real and coeval with the core. At fixed stellar
  mass, the rotation rate and
  lithium abundance are strongly anti-correlated. And finally, given Gaia-selected halo
  members, the contamination rate of field stars in the halo (XX\%) is
  larger than that in the core (YY\%).  This work expands the set of
  confirmed NGC\,2516 members by a factor of $\approx$2, and
  quantifies the degree to which Gaia-based analyses can 
  identify dispersed stellar populations against a background of field
  star contaminants.  We highlight the implications for spectroscopic
  survey targeting, open cluster dispersal, and planet searches around
  young stars.
\end{abstract}

@@

Recent clustering studies using Gaia have begun to report the
identification of structures that could correspond to low-density halos
of stars that may have evaporated from open clusters (e.g., Kounkel+19,
Kounkel+20, Meingast+21).  However, different clustering methods on the
Gaia data tend to give different results (CITE: Hunt \& Reffert 2020).
Employing say Gaussian Mixture Modelling, or any analogous method that
requires ``clusters'' to be ellipses in phase-space (position or
velocity) unsurprisingly yields open clusters that are roughly elliptical.
Some unsupervised clustering methods lead to comparable ``roughly
elliptical'' results \citep[][hereafter
\citetalias{cantatgaudin_gaia_2018}]{cantatgaudin_gaia_2018}.  Other
unsupervised methods such as HDBScan (e.g., \citep[][hereafter
\citetalias{kounkel_untangling_2019}]{kounkel_untangling_2019}) have
been found to yield additional structures, particularly in lower density
regions such as the Psc-Eri stream (CITE: Meingast+18, Curtis+19,
Newton+21, but reword bc discovery didn't use HDBScan), but also more
generally around many open clusters \citep{kounkel_untangling_2019}.
Unsupervised approaches that incorporate physical constraints like
imposing a maximum velocity dispersion on putative members yield similar
results, potentially with higher purity (Meingast+21).

@@

As part of a broader project of identifying a large and clean sample of
young stars for a transit search, we focus in this paper on a 
modest question: in just a single rich southern open cluster, is the
cluster halo coeval with the core?  The cluster we chose for this
analysis was NGC\,2516, since it is young ($\sim$120\,Myr) and close
($d\approx400$\,pc) enough to facilitate rotation measurements using
TESS, and some degree of spectroscopic analysis.  We want to know: is
the halo real? To what extent can we use Gaia alone to reliably identify
age-dated needles in the haystack of boring field stars?  And more
generally, what are the implications for the evolution of open clusters
if they do have halos?

@@

We chose the stars to focus on based on previously reported
memberships of NGC\,2516, available in the literature.  While a number
of pre-Gaia catalogs were available, the purity and accuracy of the
Gaia-derived results are expected to be the current state of the art.
We therefore adopted what we viewed as the most interesting clustering
samples to compare: those of \citet{cantatgaudin_gaia_2018}
(\citetalias{cantatgaudin_gaia_2018}) and
\citet{kounkel_untangling_2019}
(\citetalias{kounkel_untangling_2019})\footnote{A third sample was
also recently reported by \citet{meingast_2021}. Although the results
of this study were made publicly available too late to be included in
our analysis,  1577 of the 1860 \citet{meingast_2021} sources (85\%)
were included in the \citet{kounkel_untangling_2019} sample.  A
complete set of \cn\ members could very well include some of the newly
reported candidate members from \citet{meingast_2021}.}. While we
could have performed our own clustering analysis based on the Gaia
data, such an effort would be replicating the work of these
investigators. And why would our clustering methods be better than
theirs? We opt instead to use their studies as starting points.

@@

%TODO calculate these physical sizes carefully
Figure~\ref{fig:gaia6d} shows the cluster members reported by each
study.  The \citetalias{cantatgaudin_gaia_2018} members are all within
a few degrees of the cluster center, while the
\citetalias{kounkel_untangling_2019} members span tens of degrees.  In
physical units, this corresponds to a size difference of XX\,pc versus
YY\,pc.  This difference is somewhat tautological, because
\citetalias{cantatgaudin_gaia_2018} did not extend their search for
members out to tens of degrees from the cluster center, since their
clustering algorithm relied on the contrast between cluster and field
stars in the astrometric space of ($\mu_{\alpha'}, \mu_\delta, \pi$).
Nonetheless, the membership catalog of \citetalias{cantatgaudin_gaia_2018}
echoes that of many previous investigators (CITE, CITE, CITE), and is
consistent with the general {\it visual} impression that one has when
looking at \cn\ visually: it seems to be at most {\bf X$^\circ$}, and
not a big stream.
This discrepancy begs the question.  What is the true structure of
\cn?  Are the core and halo truly coeval?

To answer these questions, we adopt the
\citetalias{cantatgaudin_gaia_2018} stars as candidate ``core''
members and the \citetalias{kounkel_untangling_2019} stars as
candidate ``halo'' members.  If a star is in both lists, it is a
``core'' member.  1091 of the 1106 core members are in both lists.
This leaves 1912 candidate members in the halo.
Concatenating the two lists yields \nkinematic candidate members.

In this work, we also define a set of nearby field stars in the
``neighborhood'' of \cn.  Based on the observed distribution of halo
members, we draw these stars randomly from the following ``cube'' in
right ascension, declination, and parallax:
\begin{align}
  \alpha\,[\mathrm{deg}] &\in [108, 132], \\
  \delta\,[\mathrm{deg}] &\in [-76, -45], \\
  \pi\,[\mathrm{mas}] &\in [1.5, 4.0].
\end{align}
We imposed a magnitude limit of $G=19$\,mag, and ran the queries
using the \texttt{astroquery.gaia} module (CITE).
% TODO: why?
We allowed the number of stars in the comparison sample to exceed that
in the cluster sample by a factor of a few, to ensure broad sampling
of stellar masses and evolutionary states.
We also required the comparison sample to not overlap with the cluster
sample, which led to the omission of 1.1\% of the stars drawn over the
volume noted above.

@@

		\subfloat{
			\includegraphics[width=0.75\textwidth]{f3b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf The core and halo of \cn\ in the space of rotation
    period and Gaia color.}
    The {\it top} plot shows periods against a linear scale, while the
    {\it bottom}  shows them against a logarithmic scale.
		\label{fig:rot}
	}
\end{figure*}

@@

We then measured the rotation period of the resulting light curve.  We
used the the aperture radius that, based on theoretical expectations,
was expected to give the optimal balance between light from the target
and background-light (CITE Sullivan15).  This typically resulted in an
aperture radius of either 1 or 1.5 pixels.  To measure the periods, we
used the periodogram implementations in \texttt{astrobase}, in
particular the Stellingwerf PDM periodogram (CITE), along with the
more traditional Lomb-Scargle (CITE).  We recorded the top five
periodogram peaks from each method, and their corresponding powers.
Finally, as a check on crowding, we also recorded the number of stars
within the aperture of equal brightness to the target stars, and of
brightness with 1.25 and 2.5 TESS magnitudes of the target star.

Figure~\ref{fig:rot} shows the resulting rotation periods.  The points
on this plot are the \nautorot light curves for which the peak Lomb Scargle
periodogram period was below 15 days, the LS power exceeded 0.08, and
at most one equal-brightness companion was within the aperture.  These
selection criteria are entirely heuristic.  
But splitting the sample into the ``core'' and the ``halo'', they
suggest that the halo is real.

@@

How far away from the core, in position and velocity space, does the
halo actually extend?  We can explore this by crossmatching the
rotator sample against \nkinematic Gaia DR2 kinematic candidate members.

Figure~\ref{fig:gaia6d_x_rotn} shows the result.  An important nuance
in interpreting this plot is that we show only stars with
$0.5<(Bp-Rp)_0<1.2$ for which our TESS pipeline successfully made
detrended light curves.  In other words---the base ``core'' and
``halo'' samples are the stars for which we could have plausibly
detected a rotation period, if the star were really a cluster member.
Faint stars with $Rp>16$ were beyond our light curve selection
limit, and crowded stars ({\it e.g.,} near saturated stars
in the cluster center) do not appear.
We selected the color limit above based on Figure~\ref{fig:rot},
since the density of period detections seems to decrease once 
$(Bp-Rp)_0 \gtrsim 1.25$, {\it i.e.}, for spectral types later
than $\approx$K4V.

Qualitatively, the fraction of kinematic members with TESS-detection rotation 
periods with increasing distance from the halo core---both in
position and in velocity space.
To quantify this, we computed the distances.

@@

The third and final approach we took for age-dating the stellar
population was using the lithium deplition technique (CITE, CITE,
CITE).

Reviewing the literature, two spectroscopic datasets for \cn\ seemed
particularly important: Gaia-ESO 
\citep{randich_gaiaeso_2018} and GALAH DR3 \citep{buder_galah_2020}.
The target selection and results from each survey were as follows.

Gaia-ESO selected {\it cluster candidates} to be observed with the
GIRAFFE and UVES spectrographs based on previously reported literature
members and publicly available optical and NIR photometry.  This
target selection approach included many stars in the spatial proximity
of previously known members.  Since the possible existence of the \cn
halo was not known at the time of targetting, very few ``halo'' stars
are in their ultimate sample.  For \cn, \citet{randich_gaiaeso_2018}
ultimately reported stellar parameters (including lithium equivalent
widths and metallicities) for 796 candidate members.  Cross-matching
against our kinematic list of \nkinematic candidate members by position and
imposing a 0$\farcs$5 maximum separation limit yields 492 kinematic
members for which a Gaia-ESO spectrum was acquired and stellar
parameters were reported.  15 of these (all with $(Bp-Rp)_0 > 2.0$)
are spurious matches based on the Gaia color and GES effective
temperature, and we remove them.  This yields 477 stars, of which 436
are in the core, and 41 are in the halo.  This ratio is primarily
caused by the Gaia-ESO targetting selection function.

% GALAH DR3 selected XXX/YYY.  GALAH-DR3 = GALAH-main + GALAH-faint +
% K2-HERMES + TESS-HERMES + HERMES other.  Data taken both with the 2dF
% fiber positioner, and the HERMES spectrograph at the 3.9 AAT
% (Lewis+02, Sheinis+15).  GALAH-main is 12<V<14, $\delta<+10^\circ$,
% $|b|>10^\circ$, in regions of sky with at least 400 targets in $\pi$
% square degrees (the 2dF FoV).  Fields with more than 400 targets were
% targetted multiple times, with different fiber arrangements.
% TESS-HERMES: 10.0 < V < 13.1. TESS Southern CVZ, within $12^\circ$ of
% the southern ecliptic pole.  54354 in the ``HERMES other'' program are
% from targeted observations of stars in open clusters, the GALAH Pilot
% Survey (Martell et al. 2017),

The GALAH DR3 target selection process is discussed in detail by
\citet{buder_galah_2020}.  The most relevant aspects for our projects
are that the ``main'' survey targeted $12.0<V<14.0$ stars at
$\delta<+10^\circ$, provided they were at least ten degrees from the
galactic plane.  The survey is spatially inhomogeneous\footnote{See
the footprint at
\url{https://www.galah-survey.org/news/announcing-galah-dr3}}. Special
targetting for stars in the TESS southern continuous viewing zone, and
for known open cluster members was also performed.  We identified the
candidate \cn\ members
for which spectra had been obtained by searching
the \texttt{GALAH\_DR3\_main\_allstar\_v1} catalog, after excluding
stars with the stellar parameter bit flags 1, 2, 3.  This excludes
spectra with unreliable broadening, low S/N, and unreliable wavelength
solutions (see Table~6 of \citealt{buder_galah_2020}).  Since our main
focus is measuring equivalent widths for the Li\,6708\ \AA line, these
were the most relevant bit flags for our purpose.  Of our \nkinematic
candidate \cn\ members, 106 had spectra in GALAH DR3.  51 were in the
``core''; 55 were in the ``halo''.

%/Users/luke/Dropbox/proj/earhart/results/lithium/kinematic_X_galah_dr3.csv
% https://docs.datacentral.org.au/galah/dr3/catalogue-data-access/
We downloaded the GALAH DR3 spectra for all 106 entries by doing PROCESS X.
We then measured the lithium equivalent widths using PROCESS Y.
We then concatenated the GALAH DR3 and Gaia-ESO results together.

% TODO: fix the fact that the figure currently only shows Randich+18
Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.
The fact that we see the same trend across the core and halo of \cn
{\it a)} supports the conclusion that the halo is coeval with the
core, and {\it b)} suggests that the lithium-rotation correlation is
not explained by environmental effects such as the density of the cluster
during its embedded phase, but is instead tied to physical processes
acting on all low-mass pre-main sequence stars.

% TODO: this paragraph more or less parrots the processes described by
% Bouvier+20. are there others? is there a better way to describe
% these?
Both internal and external processes have been suggested to explain
the lithium-rotation trend (CITE see the recent review by Bouvier+20).
One explanation based in the stellar interior could be that the
convective mixing efficiency is anticorrelated with the surface
rotation (e.g., CITE Siess + Livio 1997, Baraffe+2017).  Another
possibility could be that stronger magnetic fields in the star's
interior inhibit convection (READ e.g., Ventura+98, Chabrier+07,
Somers + Pinsonneault 2014).  An external process that might also be
important is the effect of star-disk magnetic locking during the PMS
phase (CITE: magnetic braking).  Longer disk lifetimes would lead to
the star's outer convective zone being locked for longer while the
radiative core contracts.  The resulting differential rotation and
rotational mixing could drive the lithium depletion (CITE: Bouvier 08,
Eggenberger+12).
We have no preference between these possibilities---we simply note
that the \cn\ sample could be another helpful data point in
distinguishing them.

@@

\subsection{How did the halo form?}
``The stars in these halos could have escaped from the cores of the
cluster, or they could be remnants of a larger and lower-density star
formation complex that has since dispersed.''
Another possibility is {\it triggered star formation}.

\subsection{Are the ``very slow rotators'' bonafide members?}
Unsure. They are not isotropically distributed around the
cluster... so either triggered star formation (CITE, CITE), or they're
actually field stars.

\subsection{What is the contamination fraction in the halo? Does it change vs. location?}

\subsection{Mass differences between center and outer reaches?}

\subsection{Fast and blue rotators: are we going faster than other clusters?}

@@

We analyzed X, Y, Z.
Our main results are as follows.
\begin{itemize}
	\item 
\end{itemize}

@@

%
% TODO: the following two paragraphs, when quoted, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citetalias{cantatgaudin_gaia_2018} applied a procedure that 
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by DAML/Dias (CITE).  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.
When we selected candidate \cn\ members from the results of
\citetalias{cantatgaudin_gaia_2018}, we opted to include all candidate members
with reported membership probability exceeding 10\%.
While this seems {\it a priori} low, our results ({\bf SECTION XXX})
show that this ``membership probabilty'' severely underestimates the
purity of the \citetalias{cantatgaudin_gaia_2018} sample for \cn. 
%TODO: quantify
Their false positive rate across the sample is more like 1-5\%.

\citetalias{kounkel_untangling_2019}
applied a different unsupervised
clustering method to the 5-dimensional Gaia DR2 data (omitting radial
velocities, due to their sparsity).  Their selection function (see
their Section~2) yielded $\approx 2\times 10^7$ stars, mostly within
$\approx 1$\,kpc and typically with $G<18$\,mag.  Their clustering
algorithm, which was run on this entire  stellar sample,
 was the ``hierarchical density-based spatial clustering of
applications with noise'' (HDBSCAN, CITE McInnes17).  The classical
DBSCAN algorithm ``identifies clusters as overdensities in a
multi-dimensional space in which the number of sources exceeds the
required minimum number of points within a neighborhood of a
particular linking length $\epsilon$.  HDBSCAN does not depend on
$\epsilon$; instead it condenses the minimum spanning tree by pruning
off the nodes that do not meet the minimum number of sources in a
cluster and reanalyzing the nodes that do. Depending on the chosen
algorithm, it would then either find the most persistent structure
(through the excess of mass method), or return clusters as the leaves
of the tree (which results in somewhat more homogeneous clusters). In
both cases it is more effective at finding structures of varying
densities in a given data set than DBSCAN.''
`` The two main parameters that control HDBSCAN are the number of
sources in a cluster and the number of samples. The former is the
parameter that rejects groupings that are too small; the latter sets
the threshold of how conservative the algorithm is in its
considerations of the background noise (even if the resulting noisy
groupings do meet the minimum cluster size).  By default, the sample
size is set to the same value as the cluster size, but it is possible
to adjust them separately.''
Regarding membership proabilities,
\citetalias{kounkel_untangling_2019} did not report continuous membership
probabilities, instead opting for the binary ``member'' or ``not''.

@@

To determine the coefficients of the linear
model after the full set of eigenvectors (plus optionally
``sytematic'' vectors) had been asssembled,  we explored two possible
methods: ordinary
least squares, and ridge regression. Ridge regression is the same as
ordinary least squares, except it includes an L2 norm with a regularization
coefficient. The regularization coefficient that best applied for any
given target light curve was solved for using a cross-validation grid
search, using \texttt{sklearn.linear\_modelRidgeCV} (CITE).

Each target light curve was mean-subtracted and normalized by its
standard deviation, as were the eigenvectors. The linear problem was
then solved numerically, and the light curve was reconstructed by re-adding the
original mean, and re-multiplying by the standard deviation to ensure
that the variance of the light curve did not change.

We found that the choice of using ordinary
least squares versus ridge regression did not seem to significantly
affect the resulting light curves. In other words, the inclusion (or
lack thereof) of a regularization term did not strongly alter the
best-fitting coefficients.
In the spirit of ``KISS'', we opted to use ordinary least squares.

@@

Figure~\ref{fig:rotn_X_RUWE}.

\section{Kinematics $\otimes$ Rotation in EDR3}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f7.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Gaia-based components of NGC\,2516 in position and
    velocity space, cross-matched against the rotators. Analog of
    Figure~\ref{fig:gaia6d_x_rotn}, but showing EDR3 kinematics.}
		\label{fig:gaia6d_x_rotn_EDR3}
	}
\end{figure*}

How did we do?
Figure~\ref{fig:gaia6d_x_rotn_EDR3} lets us check with a more precise
astrometric solution.
